---
title: "220598721"
output: html_document
date: "03-04-2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ST2195 Programming for Data Science Coursework

## Part 2
Year 1996 to year 2000 were chosen as the consecutive five subset years in this part.
The dataset gone through exploratory data analysis process to ensure the completeness of the dataset.

```{r Data-preprocessing}
# Load all the necessary packages 
library(tidyverse)
library(lubridate)
library(ggplot2)
library(broom)
library(DataExplorer)
library(corrplot)

# Load datasets from 1996 to 2000 manually
flights_1996 <- read.csv("1996.csv")
flights_1997 <- read.csv("1997.csv")
flights_1998 <- read.csv("1998.csv")
flights_1999 <- read.csv("1999.csv")
flights_2000 <- read.csv("2000.csv")

flights_raw <- bind_rows(flights_1996, flights_1997, flights_1998, flights_1999, flights_2000)

# Exploratory Data Analysis (EDA) on the dataset
# Sample the first 5000 rows of data since it is a big dataset
flights_eda <- flights_raw %>% sample_n(min(5000, nrow(flights_raw)))

## Check structure of dataset
str(flights_eda)

## Get summary statistics for numeric columns
summary(select_if(flights_eda, is.numeric))

## Count missing values in each column
colSums(is.na(flights_eda))

## Visualize missing values
plot_missing(flights_eda)

## Show percentage of missing values per column
missing_percent <- colMeans(is.na(flights_eda)) * 100
print(missing_percent[missing_percent > 0])  # Only columns with missing values

## Name the columns with 100% missing values
cols_all_na <- names(flights_eda)[colMeans(is.na(flights_raw)) == 1]

## Remove rows with NA, ignoring the 100% missing columns
cleaned_flights <- flights_eda %>%
  select(-all_of(cols_all_na)) %>%  # Remove 100% missing columns
  filter(if_all(everything(), ~ !is.na(.)))  # Remove rows with any NA

## Display structure of cleaned dataset
str(cleaned_flights)

# Compute correlation matrix for numeric columns
corr_matrix <- cor(numeric_cols, use = "pairwise.complete.obs")

## Visualize correlation (adjust text size for large datasets)
corrplot(corr_matrix, method = "color", tl.cex = 0.6)
```

## a) The best times and days to minimise the flights delays.
```{r Question-a}
## Convert dates and times
flights <- cleaned_flights %>% 
  mutate(DATE = make_date(Year, Month, DayofMonth),
         DAY_OF_WEEK = factor(DayOfWeek, 
                              levels = 1:7, 
                              labels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
         YEAR = Year)

flights_1 <- flights %>% mutate(TOTAL_DELAY = ArrDelay + DepDelay)

## Average delay by hour
hourly_delays <- flights_1 %>%
  group_by(YEAR, HOUR = floor(CRSDepTime / 100)) %>%
  summarise(Mean_Delay = mean(TOTAL_DELAY, na.rm = TRUE))

## Find the minimum delay by hour for each year
min_delays <- hourly_delays %>% group_by(YEAR) %>% slice_min(Mean_Delay, n = 1)

## Plot delays by hour
ggplot(hourly_delays, aes(x = HOUR, y = Mean_Delay, color = as.factor(YEAR))) +
  geom_line() +
  geom_point(data = min_delays, aes(x = HOUR, y = Mean_Delay), size = 3) +
  geom_text(data = min_delays, aes(x = HOUR, y = Mean_Delay, label = round(Mean_Delay, 1)), color = "black", vjust = -0.5) +
  scale_x_continuous(breaks = seq(0, 24, by = 2), limits = c(0, 24), labels = sprintf("%02d00", seq(0, 24, by = 2))) +
  labs(title = "Average Flight Delay by Hour", x = "Scheduled Departure Hour (24-hour format)", y = "Mean Delay (mins)")

## Average delay by day
daily_delays <- flights_1 %>%
  group_by(YEAR, DAY_OF_WEEK) %>%
  summarise(Mean_Delay = mean(TOTAL_DELAY, na.rm = TRUE))

## Find the minimum delay for each year across the week
min_weekly_delays <- daily_delays %>% group_by(YEAR) %>% slice_min(Mean_Delay, n = 1)

## Plot delays by day
ggplot(daily_delays, aes(x = DAY_OF_WEEK, y = Mean_Delay, fill = as.factor(YEAR))) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(data = min_weekly_delays, aes(x = DAY_OF_WEEK, y = Mean_Delay, label = round(Mean_Delay, 1)), color = "black", vjust = -0.5, position = position_dodge(width = 0.9)) +
  labs(title = "Average Flight Delay by Day of the Week", x = "Day of Week", y = "Mean Delay (mins)")
```

## b) Evaluate whether older planes will suffer more flight delays on a year-to-year basis.
```{r Question-b}
planes_raw <- read.csv("plane-data.csv") 
planes <- planes_raw %>% rename(TailNum = tailnum)
flights_raw2 <- flights %>% mutate(TOTAL_DELAY = ArrDelay + DepDelay)
flights_2 <- left_join(flights_raw2, planes, by = "TailNum")

# Group by year and age, then calculate mean delay
age_delays <- flights_2 %>%
  mutate(AGE = 2025 - as.numeric(Year)) %>%
  group_by(YEAR, AGE) %>%
  summarise(Mean_Delay = mean(TOTAL_DELAY, na.rm = TRUE))

# Create an Age group categorization
age_delays <- age_delays %>%
  mutate(Age_Group = case_when(
    AGE <= 10 ~ "Young (0-10 years)",
    AGE > 10 & AGE <= 20 ~ "Medium (11-20 years)",
    AGE > 20 ~ "Old (>20 years)"
  ))

# Plot delays by age group for each year
ggplot(age_delays_filtered, aes(x = AGE, y = Mean_Delay, color = Age_Group)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Flight Delay vs. Aircraft Age by Year", 
       x = "Aircraft Age (years)", 
       y = "Mean Delay (mins)") +
  facet_wrap(~ YEAR) +  # Separate the plot by year
  theme_minimal()
```

## c) Logistic Regression for Diverted Flights across the years and the visualization of the coefficients of the features
```{r Question-c}
flights_3 <- flights_eda %>% 
  mutate(DATE = make_date(Year, Month, DayofMonth),
         DAY_OF_WEEK = factor(DayOfWeek, 
                              levels = 1:7, 
                              labels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")),
         YEAR = Year)
flights_3$Diverted <- as.factor(flights_3$Diverted)  

logistic_model <- glm(Diverted ~ CRSDepTime + CRSArrTime + Distance + UniqueCarrier + DAY_OF_WEEK , 
                      data = flights_3, family = binomial)

# Summary of model
summary(logistic_model)

# Extract and visualize coefficients
coefs <- tidy(logistic_model)
ggplot(coefs, aes(x = term, y = estimate, fill = term)) +
  geom_bar(stat = "identity") +
  labs(title = "Logistic Regression Coefficients for Flight Diversions", x = "Variable", y = "Coefficient") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# End of R code
