---
title: '220598721'
output: html_document
date: "03-04-2025"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ST2195 Programming for Data Science Coursework

## Part 1 (a)
To generate a sample consists of random numbers by using the Metropolis- Hastings algorithm, for the distribution of the probability density function as below:

\[ f(x) = \frac{1}{2} \exp(-|x|) \]

## To apply the Metropolis- Hastings algorithm:

```{r Metropolis-hastings}
set.seed(123)

Metropolis_hastings <- function(N, s, x0 = 0) {
  log_f <- function(x) { -abs(x) - log(2) } #By applying the practical tip
  x <- numeric(N)
  x[1] <- x0
  
  for (i in 2:N) {
    x_star <- rnorm(1, mean = x[i-1], sd = s)
    log_r <- log_f(x_star) - log_f(x[i-1])
    
    if (log(runif(1)) < log_r) {
      x[i] <- x_star
    } else {
      x[i] <- x[i-1]
    }
  }
  
  return(x)
}

## Set the values for all the parameters
N <- 10000
s <- 1
samples <- Metropolis_hastings(N, s)
```

## Including Histogram and Kernel Density Plots

```{r Histogram-KernelDensity, echo=TRUE}
library(ggplot2)

Part1a_df <- data.frame(samples = samples)

ggplot(Part1a_df, aes(x = samples)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50, color = "blue", fill = "blue", alpha = 0.6) +
  geom_density(color = "black", linewidth = 0.7) +
  stat_function(fun = function(x) 0.5 * exp(-abs(x)), color = "orange", linewidth = 1.2, linetype = "dashed") +
  labs(title = "Histogram and Kernel Density Plot of generated samples and estimates of f(x)",
       x = "x", y = "Density") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
 
```
## Report the Monte-Carlo estimates of the mean and standard deviation

```{r Monte-Carlo-estimates}
mean_estimate <- mean(samples)
sd_estimate <- sd(samples)

cat("Monte Carlo Estimate of Mean:", mean_estimate, "\n")
cat("Monte Carlo Estimate of Standard Deviation:", sd_estimate, "\n")
```
## Part 1 (b)
To calculate the $\widehat{R}$ and provide a plot of the values of $\widehat{R}$ across s values.

## R_hat for Different Chains

```{r R-hat-computation}
compute_Rhat <- function(N, s, J) {
  chains <- replicate(J, Metropolis_hastings(N, s, x0 = rnorm(1)), simplify = FALSE)
  
  Mj <- sapply(chains, mean)
  Vj <- sapply(chains, var)
  W <- mean(Vj)
  M <- mean(Mj)
  B <- mean((Mj - M)^2)
  
  R_hat <- sqrt((B + W) / W)
  return(R_hat)
}

N <- 2000
J <- 4
s_values <- seq(0.001, 1, length.out = 20)
Rhat_values <- sapply(s_values, function(s) compute_Rhat(N, s, J))
```

## Calculate the R-hat values with N = 2000, s = 0.001, J = 4

```{r Rhat-plot}
Rhat_at_s <- compute_Rhat(N = 2000, s = 0.001, J = 4)

cat("The R-hat value at s = 0.001 is", Rhat_at_s, "\n")

```

## Plot of R-hat values in the interval of s values between 0.001 and 1

```{r Rhat-plot}
Rhat_df <- data.frame(s_values, Rhat_values)

ggplot(Rhat_df, aes(x = s_values, y = Rhat_values)) +
  geom_line(color = "blue") +
  geom_hline(yintercept = 1.05, linetype = "dashed", color = "red") +
  ggtitle("R-hat Values Across Different s values in the interval between 0.001 and 1") +
  xlab("s") +
  ylab("R-hat") +
  theme_minimal()
```








